---
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---
### STAT 225 - Prelim Project Look Group D

```{r, include=FALSE}
require(mosaic)
require(NSM3)
require(readr)
require(dplyr)
require(Rfit)
```

Names:  Harrison Marick, Chase Henley, Joe Feldman (Group D)

What is the True Value of a College Education? 

Perhaps the most prevalent rationale for the pursuance of a college degree is the promise of economic mobility upon graduation.  However, our project aims to investigate whether this opportunity is equally available for all students.  Specifically, our group will analyze whether the racial and socio-demographic composition of a college affects the proportion of students who go on to earn higher wages after enrolling in the institution. 

### Read in the data

```{r, warning=FALSE}
data <- read.csv("https://awagaman.people.amherst.edu/stat225/datasetsS2018/groupDdata.csv")
data2<-data[,c(-6, -18, -19, -(27:34), -(42:89))] 
data2<-data2[,c(-2,-3,-5)] %>%
  as.data.frame() # Must start with a data-read in command from Prof. Wagaman
data2[,3:27]<-sapply(data2[,3:27], as.character)
data2[,3:27]<-sapply(data2[,3:27], as.numeric)
data3 <-data2 %>%
  group_by(INSTNM) %>%
  mutate(rank=min(UNITID), n=n()) %>%
  filter(n<2) %>%
  as.data.frame()
data4<-data3[,c(-1, -28, -29)]
```

### Summary command on data set

```{r}
summary(data4)
```


### Data Codebook

`PCT_WHITE`: Percent of the population from students' zip codes that is White, via Census data; quantitative


`MEDIAN_HH_INC`: Median Household Income of Students' Parents; quantitative. 


`GT_25K_P10`: Percentage of students earning at least 25k 10 years after entry; quantitative

`MD_EARN_WNE_P10`: Median income 10 years after entry of students working and not enrolled in school; quantitative.


### Analysis Plan

We plan to focus our project on explanation the relationships between `MD_EARN_WNE_P10` and various predictor variables. Below, we have highlighted a few potential predictors, their distributions, and their relationships with our primary response variable. 


#### Prelim Univariate Analysis

#####Percentage White

```{r setup, warning=FALSE, message=FALSE}
#Univariate Analysis
#favstats(data4$PCT_WHITE)
#histogram(data4$PCT_WHITE)
ggplot(data4, aes(PCT_WHITE))+geom_histogram()
#densityplot(data4$PCT_WHITE)
```

Conducting Univariate Analysis on the distribution of the percentage of students that are white, it is clear that this distribution is quite skewed left. The mean percentage is 75.509 while the median is 78.685. The data ranges from from 5.34 to 98.98 with a standard deviation of 15.879.

Intuitively, this distribution makes sense.  White people have a traditionally stronger representation in the American college system, and this distribution supports that claim as there is more density around colleges with a proportionately high white student body.

#####Median Earnings 10 Years Post College

```{r, warning=FALSE, message=FALSE}
favstats(data2$MD_EARN_WNE_P10)
ggplot(data2,aes(as.numeric(MD_EARN_WNE_P10)))+geom_histogram()+
  ggtitle("Dist Med Earnings 10 Years Post Entry")+
  xlab("Median Earnings (hundreds of dollars)")
```

The distribution of median earnings of students working and not enrolled 10 years after entry has a median of 31100dollars and a mean of 33500.42 dollars, and so we see the distribution is skewed right. The range of the distribution is 9100 to 250000 dollars, and the standard deviation is 15444.58 dollars which indicates a relatively large spread.

#####% of Students Earning At Least 25K

```{r, warning=FALSE, message=FALSE}
favstats(data4$GT_25K_P10)
ggplot(data4,aes((GT_25K_P10)))+
  geom_histogram()+ggtitle("Dist of % Students Earning 25K+")+
  xlab("% of Students Earning 25K+")
```

The distribution share of students earning over $25,000/year (threshold earnings) 10 years after entry has a median of 0.616 and a mean of 0.602, and we see the distribution is skewed left. The standard deviation is 0.175 which indicates a fairly large spread.


#####Median Household Income

As expected, the distribution of this variable is primarily bell-shaped. While there are a couple of spikes, there is one primary peak and it is relatively symmetric around a median of 57,739.61. The values for this variable range from 15,429.01 to 100,870.80. 


```{r, warning=FALSE, message=FALSE}
ggplot(data4, aes(MEDIAN_HH_INC)) +geom_histogram()
```


#### Prelim Bivariate Analysis

#####Percentage White and Median Earnings 10 Years Post College


```{r, warning=FALSE}
ggplot(data4, aes(PCT_WHITE, MD_EARN_WNE_P10)) + geom_point() + xlab("White %") +
  ylab("Median Earnings 10 Years Post College") + ggtitle("White% and Income")
```

There appears to be a weak positive linear correlation between percentage of student body that is white and median earnings 10 years after college.

```{r}
summary(rfit(MD_EARN_WNE_P10~PCT_WHITE, data=data4),overall.test="drop")
```
Above, we have fit a simple linear model using percent white to predict median earnings. Most notably, the p-value for model utility is very close to 0, indicating that there is a significant linear relationship between these two variables. This relationship is one we will continue to investigate as our project progresses.


#####Percentage White and % Student Body Earning >25k

```{r}
mod<-lm(GT_25K_P10~PCT_WHITE, data = data4)
summary(mod)
with(data4,plot(PCT_WHITE,GT_25K_P10))
abline(mod)
```

There seems to be a weak to moderate positive relationship between `PCT_WHITE` and the proportion of students earning over 25K within 10 years of graduating. 

