---
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---
### STAT 225 - Prelim Project Look Group X

```{r, include=FALSE}
require(mosaic)
require(NSM3)
require(readr)
```

Names:  (Group X)

(Tentative) Project Title:

Our project is about...

### Read in the data

```{r}
# Must start with a data-read in command from Prof. Wagaman
```

### Summary command on data set

```{r}
#summary()
```


### Data Codebook

List your variables and whether they are quantitative/qualitative (numeric vs categorical), along with other notes about the variables. Hint, to make a nice list, you need to put two spaces at the end of each line to force RMarkdown to start a new line. Check this out below:

Our variables are:  
Variable 1 -  
Variable 2 -  

### Analysis Plan

A rough outline for your proposed analysis, including univariate summaries, bivariate relationships, and the like

#### Prelim Univariate Analysis

Obtain basic univariate descriptive statistics and graphs for variables relevant to your analysis.

```{r}
#for quantitative variables, you could use
#histogram, bwplot, densityplot
#favstats
```

```{r}
# for qualitative variables, you could use
#tally
#bargraph
```

COMMENT on what you see!

#### Prelim Bivariate Analysis

Make at least one scatterplot or side-by-side boxplot to examine a bivariate relationship where the response is quantitative (needed for chapter 6, or 8-9 analysis). 

```{r}
#bwplot or xyplot
```

COMMENT on what you see!

#### Any other Analysis Thoughts


### Questions for me
