---
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---
### STAT 225 - Prelim Project Look Group X

```{r, include=FALSE}
require(mosaic)
require(NSM3)
require(readr)
```

Names:  Harrison Marick, Chase Henley, Joe Feldman (Group D)

What is the True Value of a College Education? 

Perhaps the most prevalent rationale for the pursuance of a college degree is the promise of economic mobility upon graduation.  However, our project aims to investigate whether this opportunity is equally available for all students.  Specifically, our group will analyze whether the racial and socio-demographic composition of a college affects the proportion of students who go on to earn higher wages after enrolling in the institution. 

### Read in the data

```{r}
data<-read.csv("Scorecard.csv")
data2<-data[,c(-6, -18, -19, -(27:34), -(42:89))] 
data2<-data2[,c(-1,-2,-3,-5)]#simplifying and only selecting certain columns# Must start with a data-read in command from Prof. Wagaman
```

### Summary command on data set

```{r}
summary(data2)
```


### Data Codebook

List your variables and whether they are quantitative/qualitative (numeric vs categorical), along with other notes about the variables. Hint, to make a nice list, you need to put two spaces at the end of each line to force RMarkdown to start a new line. Check this out below:


Variable 1 -  INSTM: Institution Name, qualitative
Variable 2 -

### Analysis Plan

A rough outline for your proposed analysis, including univariate summaries, bivariate relationships, and the like

#### Prelim Univariate Analysis

Obtain basic univariate descriptive statistics and graphs for variables relevant to your analysis.

```{r}
#for quantitative variables, you could use
#histogram, bwplot, densityplot
#favstats
```

```{r}
# for qualitative variables, you could use
#tally
#bargraph
```

COMMENT on what you see!


#### Prelim Bivariate Analysis

Make at least one scatterplot or side-by-side boxplot to examine a bivariate relationship where the response is quantitative (needed for chapter 6, or 8-9 analysis). 

```{r}
#bwplot or xyplot
```

COMMENT on what you see!

#### Any other Analysis Thoughts


### Questions for me


Joe: codebook for variables 1-9. One univariate analysis (proportion earning over 25k). One bivariate (% white as explanation of something)

Harrison: codebook for variables 10-18. One univariate analysis (distribution of whiteness). One bivariate (student earnings 10 years post by whiteness)

Chase: codebook for variables 19-26. One univariate analysis (median earnings 10 years post).
