---
title: "Written Paper"
author: "Chase Henley, Harrison Marick, Joe Feldman"
date: "4/29/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Abstract

###Introduction

How much money will I make after going to college? This is an important question  that undoubtedly crosses the minds of the millions of U.S. college students at some point during their education, and it also exerts an especially powerful influence on those who are debating where they want to attend school.

For many, choosing a college is one of the biggest financial decisions in life. College is getting more and more expensive every year, and people want to know which school is best for investing in an education. The return on such an investment is often measured by the economic benefits one receives years after graduation.

As students at Amherst College, an expensive private institution, it would be interesting to know the degree to which our personal investments will pay off. Of course there are benefits to a quality liberal arts education that are unrelated to money, but a primary objective of college is to qualify an individual to earn a comfortable living in his or her profession of choice. 

In our country, young adults entering college have a multitude of options when considering where they will pursue their degree, so if it is economic benefit down the road that they seek, what factors or characteristics of an U.S. institution should they consider when choosing a school?

There are a number of variables we could looked at when trying to figure out what makes a college a good investment, but coming from a school as diverse as Amherst Colleg, we decided to focus on racial demographics as our possible explanatory variable for economic success.

It is fairly well-known that white people have had a traditionally stronger representation in the American college system, and white people, on average, have also had higher incomes later in life. However, it seems this trend has recently been changing. In fact, the percentage of white students enrolled in degree-granting postsecondary education institutions during the period between 1976 and 2015 has fallen from 84% to 54% (National Center for Education Statistics). Thus, with the influx of minority students into higher education, is there a difference in the economic benefits reaped by white students versus students of color?

Specifically, our project explores differences in students' later-in-life incomes between predominantly white and non-predominantly white institutions. We measure the economic benefit derived from an institution (4-year schools only) as students' median earnings 10 years post-enrollment.

We begin our analysis by looking at basic univariate statistics and then explore variables visually using histograms and density plots— all the while using what we learned throughout the semester about bin width selection methods, kernels, and bandwidth selection methods.

Then, we perform parametric and nonparametric procedures to see if there is a difference in median incomes between students coming from predominantly white versus non-predominantly white institutions. We do a t-test for differences between groups for the parametric procedure and a rank-sum test for the nonparamtric procedure. We find concurring results from the nonparametric and parametric tests.

Moreover, we explore the bivariate relationship between the percentage of white students at a given institution and median incomes 10 years post-enrollment, and we end up building two linear models: one that is constructed parametrically with simple linear regression (SLR) and one that is constructed nonparametrically using a rank-based solution.  Our models turn out having quite similar slopes and model utilities.

Without going into too much detail on the statistical finds from these procedures, we will go ahead and say that we found evidence to conclude that students from predominantly white institutions have higher median incomes 10 years post-enrollment compared to students who went to non-prodominantly white institutions. 

However, it is our ethical obligation to consider the limitations to our project and our resulting conclusions. Our scope is constrained by the institutions we looked at in the U.S. and the observations that we were forced to remove due to missing information. With so many factors relating to financial success, there is the possibility of unaddressed confounding variables having an unforeseen effect on our conclusions as well.

Nevertheless, we can safely that we've made significant progress towards fulfilling our initial project goal, and we hope that you continue reading this paper to gain a more in-depth perspective on our findings and how they pertain to the broader question of how racial demographics of an institution relate to post-graduation economic success.

###Data

Our data was obtained from the College Scorecard, a database managed by the U.S. Department of Education, and contains 7,312 observations. Each observation is a different 4-year academic institution, and each is located within the United States. 

Data was collected by schools themselves via surveys and enrollment information, and then submitted to the U.S. Department of Education.

The two variables we looked at in our analysis are the following:

`MD_EARN_WNE_P10`-Median income ($) 10 years after entry of students working and not enrolled in school; quantitative. Used as our measure of economic benefit derived from an institution.

`PCT_WHITE`- Percent of the population from students' zip codes that is White, via Census data; quantitative. Used as a predictor for economic benefit derived from an institution.

We also created a new indicator variable, 'PredominantlyWhite', to distinguish schools that contained a study body with a percentage of white students greater than 67.09% as predominantly white from those with a percentage of white students less than or equal to 67.09%, which we declared non-predominantly white.  67.09% is the first quartile for the `PCT_WHITE`, and this cut-off was chosen somewhat arbitrarily, but with respect to the variable's distribution.


###Analysis

Before conducting any tests, either parametric or non parametric, it is important to examine the distribution of our variables of interest. 

```{r, include=FALSE, warning=FALSE}
data<-read.csv("Scorecard.csv")
data2<-data[,c(-6, -18, -19, -(27:34), -(42:89))] 
data2<-data2[,c(-2,-3,-5)] %>%
  as.data.frame() # Must start with a data-read in command from Prof. Wagaman
data2[,3:27]<-sapply(data2[,3:27], as.character)
data2[,3:27]<-sapply(data2[,3:27], as.numeric)
data3 <-data2 %>%
  group_by(INSTNM) %>%
  mutate(rank=min(UNITID), n=n()) %>%
  filter(n<2) %>%
  as.data.frame()
data4<-data3[,c(-1, -28, -29)]
```

```{r, warming=FALSE, message=FALSE}
ggplot(data4, aes(PCT_WHITE))+geom_histogram() + ggtitle("Distribution of Percentage White") + xlab("Percentage White") + ylab("Count")
```


Conducting Univariate Analysis on the distribution of the percentage of students that are white, it is clear that this distribution is quite skewed left. The mean percentage is 75.509 while the median is 78.685. The data ranges from from 5.34 to 98.98 with a standard deviation of 15.879.

Intuitively, this distribution makes sense.  White people have a traditionally stronger representation in the American college system, and this distribution supports that claim as there is more density around colleges with a proportionately high white student body.

Naturally, we explored different binwidths with our histograms in order to produce the most accurate picture of the distribution. While we explored various different binwidths, we found that Scott's and Sturge's methods of binwidth selection produced accurate histograms. In fact, R produced the same exact plot using both methods. 

```{r}
histogram(~PCT_WHITE,data=data4,breaks="Sturges",main="Sturges")
histogram(~PCT_WHITE,data=data4,breaks="Scott",main="Scott") 
```

In the interest of being as thorough as possible, we elected to also look at kernal density plots. We varied the kernal function and the bandwidth selection in order to find the ideal combination. 

After comparing different kernal functions with the default binwidth, we found the normal kernal to be the most smooth, producing what we felt was the best representation of the data. Below, we have compared different bandwidth selection methods.

```{r}
norm<-density(data4$PCT_WHITE, kernel="g",bw="nrd0",na.rm=TRUE) #saved normal kernel, default h
norm2<-density(data4$PCT_WHITE, kernel="g",bw="nrd",na.rm=TRUE)
norm3<-density(data4$PCT_WHITE, kernel="g",bw="ucv",na.rm=TRUE)
norm4<-density(data4$PCT_WHITE, kernel="g",bw="bcv",na.rm=TRUE)
norm5<-density(data4$PCT_WHITE,kernel="g",bw="SJ",na.rm=TRUE)

plot(norm$x,norm$y,ylim=c(0,0.05),type="l",xlab="PCT_WHITE", ylab="density", main="Comparing Bandwidth Selection Methods") #adjust y for scaling
lines(norm2$x,norm2$y,col=2) #red
lines(norm3$x,norm3$y,col=3) #green
lines(norm4$x,norm4$y,col=4) #blue
lines(norm5$x,norm5$y,col=5) #light blue
```

We see that the green and light blue lines are the least smooth—these were produced using the ucv and SJ bandwidth selection methods, respectively. These are both data-driven cross-validation (cv) methods. 

The black, red, and darker blue lines are all fairly similar in terms of smoothness, and these were produced with the nrd0, nrd, and ucv bandwidth selection methods. The nrd0 and nrd methods are both normal-based and work ideally with data sets that are normally distributed.` The ucv method is a cross-validation method as its name implies. Thus, considering our data is not normally distributed, we prefer the ucv method in this setting.

```{r}

```



###Results




###Conclusion




